Death By Dropdown?

### 2025-04-01 conference-materials/2025_shiny-conf/slides.qmd
Slide content updates
- Refines slide content for clarity and focus by condensing and reorganizing sections.
- Adjusts BID framework stages, removing before/after comparisons and streamlining notes.
- Updates stylesheets for consistency and corrects minor textual errors.

## BID In Action {.center background-color="#F0F0F0"}

::: {.r-fit-text}
**SoundPulse Market Research Dashboard Transformation**
:::

:::: {.columns .nonincremental .fragment}
::: {.column width="50%"}
### Before BID:
<!-- Dashboard with overwhelming filters and complex tables -->
![](img/before-bid.png){width="95%"}

- 14 dropdown filters
- 8 tabs of dense data tables
- No clear hierarchy of information
:::

::: {.column width="50%" .fragment}
### After BID:
<!-- Streamlined dashboard with clear information hierarchy -->
![](img/after-bid.png){width="95%"}

- Guided narrative pathway
- Focused insights by segment
- Progressive disclosure of details
:::
::::

::: {.notes}
**Show Impact**: Use before/after contrast to demonstrate value.

- Walk through the key issues with the "before" dashboard:
  * "Look at all those filters - users had to make decisions before seeing ANY data"
  * "No guidance on where to start - cognitive overload from the beginning"
- Highlight transformative elements in the "after" version:
  * "Notice how there's a clear path through the data"
  * "Progressive disclosure - you see what you need first, then can dig deeper"
  * "The interface itself tells a story about what matters most"
- Mention that this resulted in 43% faster decision-making (preview of results)
:::

## The BID Framework {background-color="#E6F0FF"}

::: {.r-fit-text}
A 5-Stage Behavior Insight Design Framework
:::

:::: {.columns}
::: {.column width="60%"}
![](img/bid-framework.png)
:::

::: {.column width="40%"}
1. **Notice** the Problem
2. **Interpret** the User's Need
3. **Structure** the Dashboard
4. **Anticipate** User Behavior
5. **Validate** & Empower the User
::: 
::::

@@

## Case Study: Stages 1-3 {background-color="#F9F9F9"}

:::: {.columns}
::: {.column width="50%"}
### SoundPulse's Challenges:

- **Stage 1: Notice**
    - Complex filtering overwhelmed users
    - 72% reported feeling confused
- **Stage 2: Interpret**
    - Users couldn't quickly process information
    - Key insights buried in text and numbers
- **Stage 3: Structure**
    - Related elements scattered throughout UI
    - No prioritization of important metrics
:::

::: {.column width="50%"}
### BID Principles Applied:

- **Stage 1: Notice**
    - Cognitive Load Theory
    - Hick's Law
- **Stage 2: Interpret**
    - Processing Fluency
    - Data Storytelling Framework
- **Stage 3: Structure**
    - Principle of Proximity
    - Default Effect
    - Dual-Processing Theory
:::
::::

@@

- "Let's see how SoundPulse applied the first three stages of the BID framework"
- For Stage 1 (Notice): "They identified how filtering complexity was causing cognitive overload"
- For Stage 2 (Interpret): "They discovered users needed to process key metrics instantly"
- For Stage 3 (Structure): "They needed to group related elements and establish clear visual hierarchy"
- "Now let's look at how they implemented these principles using {reactable}"
- Transition: "We'll start with a basic implementation and build up to more advanced features"
:::

@@

## {reactable} Implementation: Team Collaboration {background-color="#F9F9F9"}

```r
# Comments and annotation feature for Stage 5: Validate & Empower
columns = list(
  # Other columns as before
  comments = colDef(
    name = "Comments",
    cell = function(value, index) {
      comment_id <- paste0("comment-", index)
      tags$div(
        class = "comment-container",
        tags$textarea(
          id = comment_id,
          class = "form-control comment-box",
          rows = "2",
          placeholder = "Add a comment...",
          value = value,
          onchange = sprintf("saveComment(%s, this.value)", index)
        ),
        tags$div(
          class = "comment-buttons mt-1",
          tags$button(
            type = "button", 
            class = "btn btn-sm btn-outline-primary me-1",
            onclick = sprintf("addHighlight(%s)", index),
            "Highlight"
          ),
          tags$button(
            type = "button",
            class = "btn btn-sm btn-outline-secondary",
            onclick = sprintf("clearComment(%s)", index),
            "Clear"
          )
        )
      )
    },
    html = TRUE,
    width = 250
  )
)
```

::: {.notes}
**Collaboration Features**: Connect to later BID stages.

- "This example actually implements principles from Stage 5: Validate & Empower through collaboration"
- **WOW MOMENT #4**: "The comment and highlight functionality allows team members to collaborate directly within the dashboard"
- "This addresses the Cooperation & Coordination principle from Stage 5"
- "Notice how we're using custom HTML within reactable cells to create a mini application"
- "The onchange and onclick handlers connect to JavaScript functions for saving comments and highlighting rows"
- "This transforms the dashboard from a static display to a collaborative decision-making tool"
- "Similar features in tools like Google Sheets have transformed how teams work - we can bring that same power to Shiny"
- Transition: "Now that we've seen how to implement Stages 1-3 with reactable, let's move to Stages 4-5..."
:::

@@

## Case Study: Stages 4-5 {background-color="#F9F9F9"}

:::: {.columns}
::: {.column width="50%"}
### SoundPulse's Challenges:

- **Stage 4: Anticipate**
    - Executives comparing metrics without context
    - Teams interpreting data through confirmation bias
    - Different perspectives needed for different teams
- **Stage 5: Validate**
    - Insights lost after dashboard sessions
    - No clear path to consensus
    - Teams needed actionable summaries
:::

::: {.column width="50%"}
### BID Principles Applied:

- **Stage 4: Anticipate**
    - Anchoring Effect
    - Framing & Loss Aversion
    - Confirmation Bias
- **Stage 5: Validate**
    - Peak-End Rule
    - Cooperation & Coordination
    - Beautiful-Is-Good Stereotype
:::
::::

::: {.notes}
**Advanced Stages**: Set up context for echarts4r examples.

- "Now we'll look at how SoundPulse tackled the later stages of the BID framework"
- For Stage 4 (Anticipate): "They needed to address how cognitive biases were affecting interpretation"
- For Stage 5 (Validate): "They needed to create lasting impressions and facilitate team consensus"
- "These stages address the subtler psychological aspects of data interpretation"
- "Now let's see how they implemented these principles using {echarts4r}"
- "We'll start with basic chart implementation and build up to more interactive features"
:::

@@

:::: {.columns}
::: {.column width="60%"}
`{bidux}` is our new R package currently in **Phase 1 development** that helps implement the BID framework:

- **Current Features:**
  - 📚 Concept browser: `bid_concept("processing_fluency")`
  - 🧠 Documentation functions: `bid_notice()` to `bid_validate()`

- **Coming Soon (Phases 2-4):**
  - 🛠️ UI Component Library
  - 🤖 LLM Integration 
  - 📊 Testing and Validation Tools
:::

::: {.column width="40%"}
<!-- need to either update image or replace with something like the repo README -->
![](img/bid_concept-output.png){width="95%"}
:::
::::

@@

- "We're not just sharing a framework - we're building tools to make it easy to adopt"
- "Our goal is to make these psychological principles accessible to every R developer"
- Emphasize this is open source - "We want this to become a community resource"
- Current features focus on documentation and knowledge
- Future features will include ready-to-use UI components with BID principles built in
- "The LLM integration will let you describe what you want in plain language"
- "Think of {bidux} as your behavioral science assistant for dashboard development"
- "This is in active development - we welcome contributors and collaborators"
:::

@@

# 2. Stage documentation (piped workflow)
project_docs <- bid_notice(
  problem = "Dashboard overload", 
  theory = "Cognitive Load Theory",
  evidence = "User testing results"
) |>
  bid_interpret(
    central_question = "How to simplify?",
    data_story = list(hook = "Too complex", context = "Many tabs")
  )
print(processing_fluency)
```

::: {.fragment}
More at: [github.com/bidux](https://github.com)
:::

@@

- "The concept dictionary gives you instant access to behavioral principles"
- "Each concept includes definition, code examples, and empirical evidence"
- "The stage documentation functions help structure your design thinking"
- "These functions integrate with dplyr workflows via pipes"
- "Documentation becomes part of your development process, not separate"
- Point out the commented future code to generate excitement
- "In Phase 2, you'll be able to generate UI templates with BID principles built in"
- When the link appears: "We're launching this at ShinyConf today - check out the repo"
:::

@@

:::: {.columns}
::: {.column width="50%"}
### Measurable Improvements:

- Decision time reduced by 43%
- Insights per session increased by 67%
- Stakeholder satisfaction improved from 5.2/10 to 8.7/10
- Implementation time for new metrics reduced by 30%
:::

@@

- "These aren't just theoretical improvements - SoundPulse measured real results"
- "43% faster decision-making means real business value"
- "67% more insights per session means better ROI on analytics investments"
- "Satisfaction scores jumped from mediocre to excellent"
- "30% faster implementation means developers saved time too"
- When the testimonial appears: "This kind of transformation is what BID is all about"
- "Maya's team went from dreading dashboard development to using it as a competitive advantage"
- "Remember, this wasn't just visual changes - this was a fundamental approach shift"
:::

>>>>

## BID In Action: Dashboard Transformation {.center background-color="#F0F0F0"}

::: {.fragment}
<!-- TODO: exchange for gif -->
![](img/before-bid.png)
:::

::: {.notes}
**Show Impact**: Build up intuition of annoyance/frustration using "before" dashboard.

- Walk through the key issues with the "before" dashboard:
  * "Look at all those filters - users had to make decisions before seeing ANY data"
  * 8 tabs of dense data tables...need to know what you want to figure out in advance
  * "No guidance on where to start - cognitive overload from the beginning"
  * No one wants to use a dashboard like this...and no one will...UIs like this are a major contributor to dashboard rot
:::

## Behavior Insight Design (BID) Framework {background-color="#E6F0FF"}

![](img/bid-framework.png)

:::: {.columns}
::: {.column width="45%"}
1. **Notice** the Problem
2. **Interpret** the User's Need
3. **Structure** the Dashboard
:::

::: {.column width="45%"}
4. **Anticipate** User Behavior
5. **Validate** & Empower the User
::: 
::::

@@

## SoundPulse Case Study: BID Stages 1-3 {background-color="#F9F9F9"}

:::: {.columns}
::: {.column width="50%"}
- **Stage 1: Notice**
  * Complex filtering overwhelmed users (Cognitive Load)
  * 72% users reported confusion (Hick's Law)

- **Stage 2: Interpret**
  * Intuitive visualizations (Processing Fluency)
  * Clearer key insights (Data Storytelling)
:::

::: {.column width="50%"}
- **Stage 3: Structure**
  * Group related elements (Principle of Proximity)
  * Prioritization of metrics (Dual-Processing Theory)

<!-- TODO: consider adding a visual? otherwise, clean up text more -->
<!-- ![](img/small-dashboard-snippet.png){width="80%"} -->
:::
::::

@@

- "Let's see how we assessed some of the challenges SoundPulse faced and how we applied the first 3 stages of the BID framework to address them"
- For Stage 1 (Notice): "They identified how filtering complexity was causing cognitive overload"
- For Stage 2 (Interpret): "They discovered users needed to process key metrics instantly"
- For Stage 3 (Structure): "They needed to group related elements and establish clear visual hierarchy"
- "Now let's look at how we implemented these principles using {reactable}..."
:::

@@

## SoundPulse Case Study: BID Stages 4-5 {background-color="#F9F9F9"}

:::: {.columns}
::: {.column width="50%"}

- **Stage 4: Anticipate**
  * Executives comparing metrics without context (Anchoring Effect)
  * Teams interpreting data differently (Confirmation Bias)
  * Different perspectives needed for different teams (Framing & Loss Aversion)
:::

::: {.column width="50%"}
- **Stage 5: Validate**
  * Teams needed actionable summaries (Peak-End Rule)
  * No clear path to consensus (Cooperation & Coordination)

<!-- TODO: consider adding a visual? otherwise, reorganize text -->
<!-- ![](img/small-dashboard-snippet2.png){width="80%"} -->
:::
::::

::: {.notes}
**Advanced Stages Case Study**: Set up context for echarts4r examples.

- "Now we'll look at how we tackled the later stages of the BID framework in SoundPulse's dashboard..."
- For Stage 4 (Anticipate): "We needed to address how cognitive biases were affecting their interpretations of the data"
- For Stage 5 (Validate): "We needed to create lasting impressions and facilitate team consensus"
- "These stages address the subtler psychological aspects of data interpretation"
- "Now let's see how we implemented these principles using {echarts4r}"
:::

@@

@@

:::: {.columns}
::: {.column width="48%"}

- **Current Features (Phase 1):**
  * 📚 Concept browser
    * `bid_concept("processing fluency")`
  * 🧠 BID stage functions
    * `bid_notice()` to `bid_validate()`
:::

::: {.column width="48%"}

- **Coming Soon (Phases 2-4):**
  * 🤖 LLM Integration 
  * 🛠️ UI Component Library
  * 📊 Testing and Validation Tools
:::
::::

@@

- "We're not just sharing a framework - we're building tools to make it easy to adopt"
- "We're excited to introduce `{bidux}`...our new R package currently in **Phase 1 development** that helps implement the BID framework with your workflows"
- "Our goal is to make these psychological principles accessible to every R developer"
- Current features focus on documentation and knowledge
- Future features will include ready-to-use UI components with BID principles built in
- "The LLM integration will let you describe what you want in plain language"
- "Think of {bidux} as your behavioral science assistant for dashboard development"
- "This is in active development - and not quite stable yet, but please stay tuned"
:::

@@

# 2. Stage documentation (piped workflow)
project_docs <- bid_notice(
  problem = "Dashboard overload", 
  theory = "Cognitive Load Theory",
  evidence = "User testing results"
) |>
  bid_interpret(
    central_question = "How to simplify?",
    data_story = list(hook = "Too complex", context = "Many tabs")
  )
print(project_docs)
```

::: {.fragment}
More at: [github.com/jrwinget/bidux](https://github.com/jrwinget/bidux)
:::

@@

- "The concept dictionary gives you instant access to behavioral principles"
- "Each concept includes definition, code examples, and empirical evidence"
- "The stage documentation functions help structure your design thinking"
- "These functions integrate with dplyr workflows via pipes"
- "Documentation becomes part of your development process, not separate"
- Point out the commented future code to generate excitement
- "In Phase 2, you'll be able to ..."
- When the link appears: "We're launching this at ShinyConf today - check out the repo"
:::

@@

:::: {.columns}
::: {.column width="50%"}
### Real-World Impact:

- Dramatically faster decision-making
- Substantial increase in insights per session
- Stakeholder satisfaction transformed from frustration to enthusiasm
- Significantly streamlined implementation of new metrics
:::

@@

- These outcomes reflect consistent feedback across client organizations that have embraced our framework. The qualitative data tells a compelling story:
  * (PAUSE) Teams are making decisions with much greater speed and confidence
  * (PAUSE) Users are discovering more actionable insights in each session
  * (PAUSE) And perhaps most telling—stakeholders have shifted from being skeptical to becoming advocates
- We're currently building telemetry into our dashboards to gather more precise measurements...but the message from our users is already clear: this framework is changing how teams interact with their data
- When the testimonial appears: "This kind of transformation is what BID is all about"
- "Remember, this wasn't just visual changes - this was a fundamental shift in how our users...and developers...approached the dashboard"
:::
